---
title: "Biomarkers of ASD"
subtitle: "If you want a subtitle put it here"
author: "List names here"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. In all, your report can be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
library(knitr)
```

## Abstract

Write a brief one-paragraph abstract that describes the contents of your write-up.

## Dataset

A total of 154 pediatric subjects were enrolled in the study, consisting of 76 boys with autism spectrum disorder (ASD; mean age 5.6 years, SD 1.7) and 78 typically developing (TD) boys (mean age 5.7 years, SD 2.0). The ethnic distribution was 73 White/Caucasian, 32 Hispanic/Latino, 17 African American/Black, 5 Asian or Pacific Islander, 23 of multiple or other ethnicities, and 4 not reported. Autism Diagnostic Observation Schedule (ADOS) scores were also collected for all participants.

Serum protein expression was measured using the SOMAscan® assay, which quantified 1,317 proteins in 150 μl serum from 154 participants. An additional 14 blinded duplicate samples (7 ASD, 7 TD) were run for quality control but not included in analyses. In the published study, 192 proteins were removed during quality control, leaving 1,125 for analysis. The dataset used here retains all 1,317 raw protein measurements prior to filtering. These proteins span a broad range of functional classes, including cytokines, growth factors, receptors, and signaling molecules. Protein levels were reported as relative fluorescence units (RFUs), reflecting their relative abundance.

Preprocessing for the present analysis involved three main steps applied to all protein measures: a log10 transform to stabilize variance and reduce skew, z-score standardization (mean 0, SD 1) to place proteins on a comparable scale, and trimming at ±3 SD to cap extreme outliers. After these transformations, the dataset was organized with group and ados leading, followed by the processed protein features, and saved in R binary format (`biomarker_clean`) for downstream statistical analysis.

## Summary of published analysis

Summarize the methodology of the paper in 1-3 paragraphs. You need not explain the methods in depth as we did in class; just indicate what methods were used and how they were combined. If possible, include a diagram that depicts the methodological design. (Quarto has support for [GraphViz and Mermaid flowcharts](https://quarto.org/docs/authoring/diagrams.html).) Provide key results: the proteins selected for the classifier and the estimated accuracy.

## Findings

Summarize your findings here. I've included some subheaders in a way that seems natural to me; you can structure this section however you like.

### Impact of preprocessing

**TODO**


### Outliers Analysis

Outlier analysis of per-subject counts (\|z\| \> 3) showed that ASD and TD groups had similar medians (both 8.5; Wilcoxon p = 0.82), but TD participants exhibited greater dispersion with higher means, larger maxima, and a heavier right tail. Using Tukey’s 1.5×IQR rule, 9 of 78 TD participants (11.5%) were flagged as outliers compared to 4 of 76 ASD participants (5.3%). Overall, while the central tendency of outlier counts was similar across groups, extreme outlying subjects were more common in the TD group.

```{r}
#| echo: false

# -------------------
# Read headers
# -------------------
var_names <- read_csv('../data/biomarker-raw.csv',
                      col_names = FALSE,
                      n_max = 2,
                      col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, abbreviation = V2) %>%
  na.omit()

# -------------------
# Load data (no trimming); log10 + z-score
# -------------------
biomarker_no_trim <- read_csv('../data/biomarker-raw.csv',
                              skip = 2,
                              col_select = -2L,
                              col_names = c('group', 'empty', pull(var_names, abbreviation), 'ados'),
                              na = c('-', '')
) %>%
  filter(!is.na(group)) %>%
  mutate(across(-c(group, ados),
                ~ scale(log10(.x))[, 1])) %>%
  select(group, ados, everything())
```

**Group summaries.**\

We computed a per-subject outlier count as the number of proteins with \(|z|>3\) after log10 transform and z-scoring. The first table summarizes these counts by group (ASD vs TD), reporting \(N\), mean, median, SD, and range. The second table shows how often subjects exceed selected thresholds (≥10, 25, 50, 100, 200 outliers). Together, these summaries indicate the central tendency (medians) and overall dispersion, while the threshold proportions highlight the prevalence of extreme cases. If medians are similar but one group shows a higher mean, larger SD, and larger maxima with higher proportions above the thresholds, that pattern suggests a heavier right tail (i.e., more subjects with unusually many outliers) in that group.


```{r}
# ------------------
# Per-subject outlier count (|z| > 3)
# -------------------
outlier_counts <- biomarker_no_trim %>%
  mutate(outlier_n = rowSums(across(-c(group, ados), ~ abs(.x) > 3), na.rm = TRUE)) %>%
  select(group, ados, outlier_n)

# -------------------
# Group summary (ASD vs TD)
# -------------------
summary_tbl <- outlier_counts %>%
  group_by(group) %>%
  summarise(
    N      = n(),
    Mean   = mean(outlier_n),
    Median = median(outlier_n),
    SD     = sd(outlier_n),
    Min    = min(outlier_n),
    Max    = max(outlier_n),
    .groups = "drop"
  ) %>%
  arrange(group) %>%
  mutate(across(c(Mean, Median, SD), ~ round(.x, 2)))

# -------------------
# Proportions above thresholds
# -------------------
thresholds <- c(10, 25, 50, 100, 200)
props_tbl <- map_dfr(
  thresholds,
  ~ outlier_counts %>%
    group_by(group) %>%
    summarise(
      threshold       = .x,
      num_at_or_above = sum(outlier_n >= .x),
      N               = n(),
      proportion      = num_at_or_above / N,
      .groups = "drop"
    )
) %>%
  arrange(threshold, group) %>%
  mutate(proportion = round(proportion, 3))

# Show tables
kable(summary_tbl,
            caption = "Per-subject outlier counts (|z| > 3) after log10 transform and z-scoring (no trimming)")

kable(props_tbl,
            caption = "Proportion of subjects with outlier counts at or above selected thresholds (|z| > 3)")
```  
&nbsp;

**Descriptive distributions (histogram & boxplot).**\

The histogram of per-subject outlier counts shows that both ASD and TD groups cluster around similar values, but the TD group has a heavier right tail, with more subjects showing very high counts. The boxplot confirms this: medians are identical across groups, but TD participants show greater spread and more extreme high values. Tukey-flagged outlier participants are visible as red points in the boxplot.


```{r}
#| echo: false

# -------------------
# Flag subject-level outliers (Tukey rule per group)
# -------------------
fences <- outlier_counts %>%
  group_by(group) %>%
  summarise(
    Q1  = quantile(outlier_n, 0.25, na.rm = TRUE),
    Q3  = quantile(outlier_n, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    lower = Q1 - 1.5 * IQR,
    upper = Q3 + 1.5 * IQR,
    .groups = "drop"
  )

outlier_flagged <- outlier_counts %>%
  left_join(fences, by = "group") %>%
  mutate(is_outlier = outlier_n < lower | outlier_n > upper)

# -------------------
# Boxplot with outliers highlighted
# -------------------
p_box <- ggplot(outlier_flagged, aes(x = group, y = outlier_n, color = is_outlier)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "gray40", "TRUE" = "red")) +
  labs(
    title = "Per-Subject Outlier Counts by Group",
    x = "Group",
    y = "Number of Outliers (|z| > 3)",
    color = "Flagged as outlier participants"
  ) +
  theme_minimal()

p_box

# -------------------
# Histogram by group (faceted)
# -------------------
p_hist <- ggplot(outlier_counts, aes(x = outlier_n, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.6) +
  facet_wrap(~ group, ncol = 1, scales = "free_y") +
  labs(
    title = "Histogram of Per-Subject Outlier Counts",
    x = "Number of Outliers (|z| > 3)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p_hist
```  
&nbsp;

**Wilcoxon rank-sum test.**\
The test of group differences in outlier counts showed no evidence of a location shift (W = 3028.5, p = 0.817), consistent with identical medians (8.5).

```{r}
#| echo: false

# --- Wilcoxon rank-sum (location) ---
wilcox_res <- wilcox.test(outlier_n ~ group, data = outlier_counts)
wilcox_tbl <- tibble::tibble(
  test    = "Wilcoxon rank-sum",
  W       = unname(wilcox_res$statistic),
  p_value = wilcox_res$p.value
)
knitr::kable(wilcox_tbl, caption = "Wilcoxon rank-sum test comparing per-subject outlier counts (ASD vs TD)")

# --- Flagged subjects per group (Tukey rule) ---
outlier_totals <- outlier_flagged %>%
  dplyr::count(group, is_outlier) %>%
  tidyr::complete(group, is_outlier, fill = list(n = 0)) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

knitr::kable(outlier_totals,
             caption = "Flagged outlier participants by group (Tukey 1.5×IQR rule) with proportions")

# --- 2×2 table + chi-square test for flagged proportions ---
tab <- table(outlier_flagged$group, outlier_flagged$is_outlier)
chisq_res <- chisq.test(tab)

# format the contingency table for display
tab_df <- as.data.frame.matrix(tab) %>%
  tibble::rownames_to_column("Group")
knitr::kable(tab_df, caption = "2×2 contingency table: flagged (TRUE/FALSE) by group")

chisq_tbl <- tibble::tibble(
  test    = "Chi-square test of independence",
  statistic = unname(chisq_res$statistic),
  df        = unname(chisq_res$parameter),
  p_value   = chisq_res$p.value
)
knitr::kable(chisq_tbl, caption = "Chi-square test comparing flagged proportions across groups")
```  
&nbsp;


**Overall interpretation.**\
- Both groups share a similar central tendency (same medians, Wilcoxon not significant).\
- The TD group shows greater dispersion and heavier tails, including more extreme subjects.\
- Using Tukey’s rule, a larger proportion of TD participants were classified as outliers compared with ASD.

**Outlier Analysis Conclusion:** While typical participants look similar in ASD and TD, extremely outlying individuals are more common in the TD group.


### Methodological variations

Task 3

### Improved classifier

Task 4
