---
title: "Biomarkers of ASD"
subtitle: "Evaluating the impact of preprocessing, outlier detection, and selection strategies on ASD biomarker accuracy"
author: "Jeff Loomis, Sanil Katula"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. In all, your report can be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
library(knitr)
```

## Abstract

Write a brief one-paragraph abstract that describes the contents of your write-up.

## Dataset

A total of 154 pediatric subjects were enrolled in the study, consisting of 76 boys with autism spectrum disorder (ASD; mean age 5.6 years, SD 1.7) and 78 typically developing (TD) boys (mean age 5.7 years, SD 2.0). The ethnic distribution was 73 White/Caucasian, 32 Hispanic/Latino, 17 African American/Black, 5 Asian or Pacific Islander, 23 of multiple or other ethnicities, and 4 not reported. Autism Diagnostic Observation Schedule (ADOS) scores were also collected for all participants.

Serum protein expression was measured using the SOMAscan® assay, which quantified 1,317 proteins in 150 μl serum from 154 participants. An additional 14 blinded duplicate samples (7 ASD, 7 TD) were run for quality control but not included in analyses. In the published study, 192 proteins were removed during quality control, leaving 1,125 for analysis. The dataset used here retains all 1,317 raw protein measurements prior to filtering. These proteins span a broad range of functional classes, including cytokines, growth factors, receptors, and signaling molecules. Protein levels were reported as relative fluorescence units (RFUs), reflecting their relative abundance.

Preprocessing for the present analysis involved three main steps applied to all protein measures: a log10 transform to stabilize variance and reduce skew, z-score standardization (mean 0, SD 1) to place proteins on a comparable scale, and trimming at ±3 SD to cap extreme outliers. After these transformations, the dataset was organized with group and ados leading, followed by the processed protein features, and saved in R binary format (`biomarker_clean`) for downstream statistical analysis.

## Summary of published analysis

The study by Hewitson et al. (2021) aimed to identify serum protein biomarkers that could distinguish children with autism spectrum disorder (ASD) from typically developing (TD) controls. Blood samples from 154 male children (76 ASD and 78 TD) were analyzed using the SOMAscan® 1.3K proteomic platform, which quantified 1,317 serum proteins. After quality control procedures excluded 192 proteins, the remaining 1,125 were log transformed, standardized to z-scores, and values beyond ±3 standard deviations were trimmed to reduce the influence of extreme outliers.

Three feature selection approaches were applied: two-sample t-tests comparing ASD and TD groups, random forest importance rankings, and Spearman correlations between protein levels and ADOS scores. From each method, the ten most informative proteins were selected, and the overlapping set across all three approaches consisted of IgD, suPAR, MAPK14, EPHB2, and DERM, which defined the core panel.

Logistic regression models were then used to evaluate classification performance and to assess whether adding additional proteins could improve accuracy. Four additional proteins, ROR1, GI24, eIF-4H, and ARSB, each improved classification performance when added to the five-protein core set. The combined nine-protein optimal panel achieved an AUC of 86% (sensitivity 83%, specificity 84%).

#### Workflow of Hewitson et al. (2021) Biomarker Analysis

```{r}
#| echo: false
if (!requireNamespace("DiagrammeR", quietly = TRUE)) install.packages("DiagrammeR")
library(DiagrammeR)

grViz("
digraph G {
  rankdir=TB;
  graph [splines=true, nodesep=0.6, ranksep=0.5];
  node  [shape=box, style=\"rounded,filled\", fillcolor=\"#F5F7FA\", color=\"#94A3B8\", fontname=\"Helvetica\"];
  edge  [color=\"#94A3B8\"];

  # Top row (data & preprocessing)
  A [label=\"Serum samples\\n154 boys (76 ASD, 78 TD)\"];
  B [label=\"SOMAscan 1.3K\\n1,317 proteins\"];
  C [label=\"Protein QC\\nremove 192\"];
  D [label=\"Retained\\n1,125 proteins\"];
  E [label=\"Preprocessing\\nlog10 transform, z-score, trim ±3 SD\"];

  # Second row (feature selection & modeling)
  M1 [label=\"t-tests\\n(ASD vs TD)\"];
  M2 [label=\"Random forest\\nimportance\"];
  M3 [label=\"Spearman correlation\\nwith ADOS\"];
  CORE [label=\"Overlap of top 10\\nfrom each method\\nIgD, suPAR, MAPK14, EPHB2, DERM\"];
  ADD4 [label=\"Add 4 proteins\\nROR1, GI24, eIF-4H, ARSB\"];
  LR [label=\"Logistic regression\\nmodeling\"];
  FINAL [label=\"Final classifier\\n9-protein panel\\nAUC 86% (sens 83%, spec 84%)\"];

  {rank=same; M1; M2; M3}

  A -> B -> C -> D -> E;
  E -> M1;
  E -> M2;
  E -> M3;

  M1 -> CORE;
  M2 -> CORE;
  M3 -> CORE;

  CORE -> ADD4 -> LR -> FINAL;
}
")
```

## Findings

Summarize your findings here. I've included some subheaders in a way that seems natural to me; you can structure this section however you like.

### Impact of preprocessing

#### Log-transformation of protein intensities

To assess the effect of log-transformation, I compared raw versus log10-transformed protein intensities.  
On the raw scale, most proteins were strongly right-skewed (median skew ≈ 2.0).  
After applying the log10 transform (with a small positive shift for non-positive values), distributions became substantially more symmetric (median skew ≈ 0.2).  
About 70 % of proteins showed reduced absolute skewness, and the relationship between the mean and standard deviation was much weaker, indicating partial variance stabilization.  

Overall, the log10 transformation reduces skewness, tames extreme high values, and stabilizes variance—important for statistical models that assume approximate normality and homoscedasticity.

```{r}
#| echo: false
#| fig-cap: "Distributions of raw (left) and log10-transformed (right) protein intensities for a random sample of nine proteins."
#| layout-ncol: 2
knitr::include_graphics(c(
  "q1_log/hist_raw_sample.png",
  "q1_log/hist_log_sample.png"
))


#| echo: false
#| fig-cap: "Q–Q plots for the same proteins before (left) and after (right) log10 transformation, showing improved approximate normality."
#| layout-ncol: 2

knitr::include_graphics(c(
"q1_log/qq_raw_sample.png",
"q1_log/qq_log_sample.png"
))

#| echo: false
#| fig-cap: "Standard deviation versus mean for each protein on the raw (left) and log10 (right) scales. The strong coupling of mean and SD on the raw scale is reduced after log-transformation."
#| layout-ncol: 2

knitr::include_graphics(c(
"q1_log/mean_sd_raw.png",
"q1_log/mean_sd_log.png"
))
```

### Outliers Analysis

Outlier analysis of per-subject counts (\|z\| \> 3) showed that ASD and TD groups had similar medians (both 8.5; Wilcoxon p = 0.82), but TD participants exhibited greater dispersion with higher means, larger maxima, and a heavier right tail. Using Tukey’s 1.5×IQR rule, 9 of 78 TD participants (11.5%) were flagged as outliers compared to 4 of 76 ASD participants (5.3%). Overall, while the central tendency of outlier counts was similar across groups, extreme outlying subjects were more common in the TD group.

```{r}
#| echo: false

# -------------------
# Read headers
# -------------------
var_names <- read_csv('../data/biomarker-raw.csv',
                      col_names = FALSE,
                      n_max = 2,
                      col_select = -(1:2)) %>%
  t() %>%
  as_tibble() %>%
  rename(name = V1, abbreviation = V2) %>%
  na.omit()

# -------------------
# Load data (no trimming); log10 + z-score
# -------------------
biomarker_no_trim <- read_csv('../data/biomarker-raw.csv',
                              skip = 2,
                              col_select = -2L,
                              col_names = c('group', 'empty', pull(var_names, abbreviation), 'ados'),
                              na = c('-', '')
) %>%
  filter(!is.na(group)) %>%
  mutate(across(-c(group, ados),
                ~ scale(log10(.x))[, 1])) %>%
  select(group, ados, everything())
```

**Group summaries**\

We computed a per-subject outlier count as the number of proteins with (\|z\|\>3) after log10 transformation and z-scoring. The first table summarizes these counts by group (ASD vs TD), reporting (N), mean, median, SD, and range. Both groups had similar medians (8.5), but TD subjects showed greater variability, with a higher mean (17.6 compared to 13.3) and a wider range of values (0–157 versus 0–126). The second table shows how often subjects exceeded selected thresholds (≥10, 25, 50, 100, and 200 outliers). Across all thresholds, a larger proportion of TD participants were above each cutoff, indicating more frequent extreme outlier counts. Together, these summaries show that while the typical number of outliers per subject was comparable between groups, the TD distribution extended further to the right, with a slightly larger number of subjects showing unusually high outlier counts.

```{r}
# ------------------
# Per-subject outlier count (|z| > 3)
# -------------------
outlier_counts <- biomarker_no_trim %>%
  mutate(outlier_n = rowSums(across(-c(group, ados), ~ abs(.x) > 3), na.rm = TRUE)) %>%
  select(group, ados, outlier_n)

# -------------------
# Group summary (ASD vs TD)
# -------------------
summary_tbl <- outlier_counts %>%
  group_by(group) %>%
  summarise(
    N      = n(),
    Mean   = mean(outlier_n),
    Median = median(outlier_n),
    SD     = sd(outlier_n),
    Min    = min(outlier_n),
    Max    = max(outlier_n),
    .groups = "drop"
  ) %>%
  arrange(group) %>%
  mutate(across(c(Mean, Median, SD), ~ round(.x, 2)))

# -------------------
# Proportions above thresholds
# -------------------
thresholds <- c(10, 25, 50, 100, 200)
props_tbl <- map_dfr(
  thresholds,
  ~ outlier_counts %>%
    group_by(group) %>%
    summarise(
      threshold       = .x,
      num_at_or_above = sum(outlier_n >= .x),
      N               = n(),
      proportion      = num_at_or_above / N,
      .groups = "drop"
    )
) %>%
  arrange(threshold, group) %>%
  mutate(proportion = round(proportion, 3))

# Show tables
kable(summary_tbl,
            caption = "Per-subject outlier counts (|z| > 3) after log10 transform and z-scoring (no trimming)")

kable(props_tbl,
            caption = "Proportion of subjects with outlier counts at or above selected thresholds (|z| > 3)")
```

 

**Descriptive distributions (histogram & boxplot)**\

The histogram of per-subject outlier counts shows that both ASD and TD groups cluster around similar values, but the TD group has a heavier right tail, with more subjects showing very high counts. The boxplot confirms this: medians are identical across groups, but TD participants show greater spread and more extreme high values. Tukey-flagged outlier participants are visible as red points in the boxplot.

```{r}
#| echo: false

# -------------------
# Flag subject-level outliers (Tukey rule per group)
# -------------------
fences <- outlier_counts %>%
  group_by(group) %>%
  summarise(
    Q1  = quantile(outlier_n, 0.25, na.rm = TRUE),
    Q3  = quantile(outlier_n, 0.75, na.rm = TRUE),
    IQR = Q3 - Q1,
    lower = Q1 - 1.5 * IQR,
    upper = Q3 + 1.5 * IQR,
    .groups = "drop"
  )

# -------------------
# Histogram by group (faceted)
# -------------------
p_hist <- ggplot(outlier_counts, aes(x = outlier_n, fill = group)) +
  geom_histogram(bins = 30, alpha = 0.6) +
  facet_wrap(~ group, ncol = 1, scales = "free_y") +
  labs(
    title = "Histogram of Per-Subject Outlier Counts",
    x = "Number of Outliers (|z| > 3)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p_hist

outlier_flagged <- outlier_counts %>%
  left_join(fences, by = "group") %>%
  mutate(is_outlier = outlier_n < lower | outlier_n > upper)

# -------------------
# Boxplot with outliers highlighted
# -------------------
p_box <- ggplot(outlier_flagged, aes(x = group, y = outlier_n, color = is_outlier)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "gray40", "TRUE" = "red")) +
  labs(
    title = "Per-Subject Outlier Counts by Group",
    x = "Group",
    y = "Number of Outliers (|z| > 3)",
    color = "Flagged as outlier participants"
  ) +
  theme_minimal()

p_box
```

 

**Group Comparisons of Outlier Counts**\

The Wilcoxon rank-sum test showed no significant difference in median outlier counts between ASD and TD participants ((W = 3028.5), (p = 0.82)). Using Tukey’s 1.5×IQR rule, 9 of 78 TD participants (11.5%) and 4 of 76 ASD participants (5.3%) were flagged as outliers. The chi-square test comparing these proportions was not significant, indicating that while extreme subjects occurred in both groups, they were slightly more common among TD participants without reaching statistical significance.

```{r}
#| echo: false

# --- Wilcoxon rank-sum (location) ---
wilcox_res <- wilcox.test(outlier_n ~ group, data = outlier_counts)
wilcox_tbl <- tibble::tibble(
  test    = "Wilcoxon rank-sum",
  W       = unname(wilcox_res$statistic),
  p_value = wilcox_res$p.value
)
knitr::kable(wilcox_tbl, caption = "Wilcoxon rank-sum test comparing per-subject outlier counts (ASD vs TD)")

# --- Flagged subjects per group (Tukey rule) ---
outlier_totals <- outlier_flagged %>%
  dplyr::count(group, is_outlier) %>%
  tidyr::complete(group, is_outlier, fill = list(n = 0)) %>%
  dplyr::group_by(group) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

knitr::kable(outlier_totals,
             caption = "Flagged outlier participants by group (Tukey 1.5×IQR rule) with proportions")

# --- 2×2 table + chi-square test for flagged proportions ---
tab <- table(outlier_flagged$group, outlier_flagged$is_outlier)
chisq_res <- chisq.test(tab)

# format the contingency table for display
tab_df <- as.data.frame.matrix(tab) %>%
  tibble::rownames_to_column("Group")
knitr::kable(tab_df, caption = "2×2 contingency table: flagged (TRUE/FALSE) by group")

chisq_tbl <- tibble::tibble(
  test    = "Chi-square test of independence",
  statistic = unname(chisq_res$statistic),
  df        = unname(chisq_res$parameter),
  p_value   = chisq_res$p.value
)
knitr::kable(chisq_tbl, caption = "Chi-square test comparing flagged proportions across groups")
```

 

**Overall interpretation**\

-   Both groups share a similar central tendency (same medians, Wilcoxon not significant).\
-   The TD group shows greater dispersion and heavier tails, including more extreme subjects.\
-   Using Tukey’s rule, a larger proportion of TD participants were classified as outliers compared with ASD.

**Outlier Analysis Conclusion:** While typical participants look similar in ASD and TD, extremely outlying individuals are more common in the TD group.

### Methodological variations

Task 3

### Improved classifier

Task 4
